---
title: "Tidy Tuesday 2025-03-18"
author: "Henry Owens"
date: "2025-03-17"
format: html
toc: true
---

# Palm Trees
From the prompt: 
*This week we're exploring Palm Trees!*

*The dataset comes from the the PalmTraits 1.0 database via the palmtrees R package by Emil Hvitfeldt.*

*Plant traits are critical to plant form and function —including growth, survival and reproduction— and therefore shape fundamental aspects of population and ecosystem dynamics as well as ecosystem services. Here, we present a global species-level compilation of key functional traits for palms (Arecaceae), a plant family with keystone importance in tropical and subtropical ecosystems.*

*- How do the sizes of the different species of palms vary across sub families?*

*- Which fruit colors occur most often?*

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(ggplot2)
tuesdata <- tidytuesdayR::tt_load('2025-03-18')
palmtrees <- tuesdata$palmtrees
```

# Data viz

This week, I did some pretty basic data visualizations for the prompt.

## How do the sizes of the different species of palms vary across sub families?

### Stem Height  
```{r}
#| echo: false
#| message: false
#| warning: false
palmtrees %>% 
  filter(max_stem_height_m > 0,
         !is.na(max_stem_height_m)) %>% 
  ggplot(aes(x = palm_subfamily, 
             y = max_stem_height_m, 
             color = palm_subfamily)) +
  geom_boxplot() +
  labs(title = "Stem Height by Subfamily",
       x = "Palm Subfamily",
       y = "Maximium Stem Height (m)",
       color = NULL) +
  guides(color = "none") +
  coord_flip()
```


### Number of Leaves
```{r}
#| echo: false
#| message: false
#| warning: false
palmtrees %>% 
  filter(max_leaf_number > 0,
         !is.na(max_leaf_number)) %>% 
  ggplot(aes(x = palm_subfamily, 
             y = max_leaf_number, 
             color = palm_subfamily)) +
  geom_boxplot() +
  labs(title = "Leaf Count by Subfamily",
       x = "Palm Subfamily",
       y = "Maximum Leaf Number",
       color = NULL) +
  guides(color = "none") +
  coord_flip()
```


### Length of Leaves/Blades
```{r}
#| echo: false
#| message: false
#| warning: false
palmtrees %>% 
  filter(max__blade__length_m > 0,
         !is.na(max__blade__length_m)) %>% 
  ggplot(aes(x = palm_subfamily, 
             y = max__blade__length_m, 
             color = palm_subfamily)) +
  geom_boxplot() +
  labs(title = "Blade Length by Subfamily",
       x = "Palm Subfamily",
       y = "Maximum Blade Length (m)",
       color = NULL) +
  guides(color = "none") +
  coord_flip()
```


## Which fruit colors occur most often?  
```{r}
#| echo: false
#| message: false
#| warning: false
### semi cool chart:
palm_plot_df <- palmtrees %>%
  filter(!is.na(main_fruit_colors)) %>%
  select(spec_name, main_fruit_colors, palm_subfamily) %>%
  separate(main_fruit_colors, into = c("dominant_color"), sep = ";", extra = "drop", remove = FALSE) %>%
  mutate(dominant_color = trimws(dominant_color)) %>% 
  group_by(palm_subfamily, dominant_color) %>%
  summarise(n_species = n_distinct(spec_name)) %>%
  mutate(subfam_portion = n_species/sum(n_species)) %>% 
  ungroup() %>%
  group_by(palm_subfamily) %>%
  arrange(desc(subfam_portion), .by_group = TRUE) %>%
  mutate(subblob_id = row_number()) %>%
  ungroup()

ggplot(palm_plot_df, aes(x = palm_subfamily, y = subblob_id, size = subfam_portion, color = dominant_color)) +
  geom_point(alpha = 0.7) +  
  scale_size_continuous(range = c(3, 20)) +  
  scale_color_manual(values = c(
    "black" = "black",
    "blue" = "blue",
    "brown" = "chocolate4",
    "green" = "darkgreen",
    "orange" = "orange",
    "pink" = "pink",
    "purple" = "purple",
    "red" = "red",
    "white" = "white",
    "yellow" = "yellow",
    "cream" = "beige",
    "grey" = "darkgrey",
    "ivory" = "ivory",
    "straw-coloured" = "wheat"
  )) +  # assign colors
  labs(
    title = "Dominant Fruit Color Frequency by Palm Subfamily",
    x = "Subfamily",
    y = NULL,
    size = "Subfamily Portion",
    color = NULL
  ) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    axis.text.y = element_blank(),  
    axis.ticks.y = element_blank(), 
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    panel.background = element_rect(fill = "lightgray")
  )
```

