---
title: "Tidy Tuesday 2025-03-04"
author: "Henry Owens"
date: "2025-03-04"
format: html
toc: true

---

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(ggmosaic)
library(gtsummary)
library(lsr)
library(chisq.posthoc.test)
library(ggplot2)
knitr::opts_chunk$set(message = FALSE, echo = FALSE, warning = FALSE)
```

## Long Beach Animal Shelter data  
The dataset comes from the City of Long Beach Animal Care Services via the {animalshelter} R package.

I did some basic dataviz and ran Chi Squared tests on a few variables. Getting back into the swing of things and buffing up on my Chi Squared assumptions and interpretations. Also this is   

https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-03-04/readme.md

## Reading in the data
```{r}
#| message: false
#| warning: false
vet_df <- readr::read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-04/longbeach.csv') %>% 
  mutate(intake_condition = str_replace(intake_condition, '  ', ' '))
print(head(vet_df))
```

## Cleaning Data
Clean up the two variables (`intake_condition`, `outcome_type`) for a Chi Squared test. Making some assumptions here, but want to reduce the number of categories.   
```{r}
# add age
vet_df <- vet_df %>% 
  filter(!outcome_type %in% c('duplicate', 'missing'), 
         !is.na(outcome_type)) %>% 
  mutate(intake_condition = case_when(str_detect(intake_condition, 'behavior moderate') ~ 'ill moderate',
                                      str_detect(intake_condition, 'behavior mild') ~ 'ill mild',
                                      str_detect(intake_condition, 'behavior severe') ~ 'ill severe',
                                      str_detect(intake_condition, 'intakeexam') ~ 'normal',
                                      TRUE ~ intake_condition
                                      ),
         intake_condition = str_replace(intake_condition, 'moderatete', 'moderate'),
         outcome_type_clean = case_when(str_detect(outcome_type,'neuter') ~ 'neuter', 
                                   str_detect(outcome_type,'return') ~ 'return', 
                                   str_detect(outcome_type, 'foster') ~ 'adoption',
                                   # lumping with adoption
                                   str_detect(outcome_type, 'homefirst') ~ 'adoption', 
                                   # not sure what this means, but lumping with adoption. could be wrong
                                   str_detect(outcome_type, 'disposal') ~ 'died', 
                                   str_detect(outcome_type, 'community cat') ~ 'return',
                                   # not sure here, but CATegory violates chisq assumption of no impossible combinations
                                   TRUE ~ outcome_type),
      age = difftime(intake_date, dob, units = 'days')/365,
      stay_length = difftime(outcome_date, intake_date, units = 'days')) 
```

## Inspecting some variables:


I want to do chi squared tests and here we look at some of the categorical variables. 
```{r}
vet_df %>%
  ggplot(aes(x = reorder(animal_type, -table(animal_type)[animal_type]))) +  # Sort in descending order
  geom_bar(fill = "darkseagreen") +  
  labs(title = "Counts of Animal Types",
       x = "Animal Type",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Bar chart for intake_condition
vet_df %>%
  ggplot(aes(x = reorder(intake_condition, -table(intake_condition)[intake_condition]))) +  # Sort in descending order
  geom_bar(fill = "darkseagreen") +  
  labs(title = "Counts of Intake Conditions",
       x = "Intake Condition",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

```{r}
# Bar chart for outcome_type
vet_df %>%
  ggplot(aes(x = reorder(outcome_type_clean, -table(outcome_type_clean)[outcome_type_clean]))) +  # Sort in descending order
  geom_bar(fill = "darkseagreen") +  
  labs(title = "Counts of Outcome Types",
       x = "Outcome Type",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```


## Chi squared: is there a relationship btw `intake_condition` & `outcome_type`?


```{r}
# Create a contingency table
contingency_table <- table(vet_df$intake_condition, vet_df$outcome_type_clean)

# Perform Chi-Square Test
chi_square_result <- chisq.test(contingency_table)
print(chi_square_result)
```

Calculate row proportions to analyze specific associations. Look for high proportions in specific cells.  
```{r}
# Calculate row proportions to analyze specific associations. Look for high proportions in specific cells.
row_proportions <- prop.table(contingency_table, margin = 1)
print(row_proportions)

```


Visualize with a heat map. You can see there are some common relationships between `intake_condition` & `outcome_type`:
- "normal" intake has a high proportion of "adoptions"
- "ill severe" and "injured severe" more associated with "euthanasia" :(  
```{r}
contingency_df <- as.data.frame(contingency_table)
ggplot(contingency_df, aes(x = Var2, y = Var1, fill = Freq)) +
  geom_tile() +
  labs(title = "Heatmap of Intake Condition vs. Outcome Type",
       x = "Outcome Type",
       y = "Intake Condition") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Post-hoc analysis: Checking associations between each specific pair of categories using the Bonferroni correction method. Positive (negative) residual means relationship is more frequent (less frequent) than expected. Values with absolute value > 2 indicate observation is significantly different from expected.  
```{r}
# Perform post-hoc tests (optional)
posthoc_results <- chisq.posthoc.test(contingency_table, method = "bonferroni")
print(posthoc_results)
```

Here we can see that severe illness is LESS associated with adoption, and MORE associated with euthanasia and death. SAD, but mild illness is more associated with adoption at least.  
```{r}
posthoc_results %>% 
  filter(str_detect(Dimension, 'ill')) %>% 
  print()
```


## Chi Squared: Checking the relationship btw `animal_type` and `outcome_type`
```{r}
# Create a contingency table
contingency_table2 <- table(vet_df$animal_type, vet_df$outcome_type_clean)

# Perform Chi-Square Test
chi_square_result2 <- chisq.test(contingency_table2)
print(chi_square_result2)
```
This is indicating a strong overall discrepancy between the observed and expected frequencies in the contingency table. P-value is very small (statistically significant). BUT we do not meet assumptions of the model since dog, cat, or bird are most of the observations. There is only 1 amphibian, e.g., so we can't possibly have 5 observations across ~10 outcomes. 

*The expected frequency in each cell of the contingency table should be at least 5 for the Chi-Square approximation to be valid.*

```{r}
contingency_df2 <- as.data.frame(contingency_table2)
ggplot(contingency_df2, aes(x = Var2, y = Var1, fill = Freq)) +
  geom_tile() +
  labs(title = "Heatmap of Intake Condition vs. Outcome Type",
       x = "Outcome Type",
       y = "Intake Condition") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Checking frequencies in contingency table. We know this will violate since there is only 1 amphibian record total, and few observations of the other animals.
```{r}
# Check expected frequencies
expected2 <- chisq.test(contingency_table2)$expected
expected2
# Check if any expected frequencies are less than 5
#any(expected2 < 5)
```

## Chi Squared: Checking the relationship btw `animal_type` (ONLY DOG, CAT, BIRD) and `outcome_type`

I don't think think this really fits the Chi squared test requirements, since the animal type and outcome are not really independent. Birds are probably pigeons and wild birds that get injured/fly into buildings are not ever going to be adopted.  
```{r}
vet_df_dc <- vet_df %>% 
  filter(animal_type %in% c('dog', 'cat', 'bird'))
# Create a contingency table
contingency_table3 <- table(vet_df_dc$animal_type, vet_df_dc$outcome_type_clean)

# Perform Chi-Square Test
chi_square_result3 <- chisq.test(contingency_table3)
print(chi_square_result3)
```

```{r}
contingency_df3 <- as.data.frame(contingency_table3)
ggplot(contingency_df3, aes(x = Var2, y = Var1, fill = Freq)) +
  geom_tile() +
  labs(title = "Heatmap of Intake Condition vs. Outcome Type",
       x = "Outcome Type",
       y = "Intake Condition") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
