---
title: "Tidy Tuesday Week 10"
author: "Henry Owens"
date: "2025-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
```

```{r message=FALSE, warning=FALSE}
# Using R
# Option 1: tidytuesdayR R package 
#install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2025-03-11')
## OR
#tuesdata <- tidytuesdayR::tt_load(2025, week = 10)

pixar_films <- tuesdata$pixar_films
public_response <- tuesdata$public_response
```
## Data Inspection
I can see the tables have different row counts, so let's investigate that.

These are in the films table, but not response table:
The NA film is Elemental, according to the release date. I assign the title because i went through the trouble of checking, but there isnt response data.  

```{r}
setdiff(pixar_films$film, public_response$film)
```

## Plotting Select Ratings

Here I make a plot comparing the Metacritic and Rotten Tomatoes scores.   
```{r}
# assign missing title
pixar_films <- pixar_films %>% 
  mutate(film = if_else(release_date == '2023-06-16', 'Elemental', film))
         

pixar_df <- merge(pixar_films, public_response, by = "film", all = TRUE) %>% 
  mutate(film_year = paste0(film, " (", year(release_date), ")"),
         film_year = fct_reorder(film_year, release_date))

pixar_df %>% 
  filter(!is.na(rotten_tomatoes), !is.na(metacritic)) %>% 
  ggplot() +
  geom_point(aes(x = film_year, y = rotten_tomatoes), color = "darkred") +
  geom_point(aes(x = film_year, y = metacritic), color = "blue") +
  # cant ge the emoji labels to work...
  # geom_text(aes(x = film_year, y = rotten_tomatoes, label = "🍅")) +
  # geom_text(aes(x = film_year, y = metacritic, label = "🎥")) +
  labs(x = "Film (year)", 
       y = "Rotten Tomatoes score (red), Metacritic score (blue)",
       title ="Comparing Different Critical Responses to Pixar Films") +
  coord_flip() +
  theme_minimal()
```





